Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
changed CFG nonterminal \sigma.

\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow
(\abigfunction\ \acdr)}$ \IN
          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow
(}$\cred{+}$\cred{\ \pa\ \acdr)}$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
$\cred{\leftarrow  (\CONS\ \pb\ ())}$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
          $\cred{(\LET\ \pw}\  $\cred{\leftarrow  (\length\ \pc)}$ \IN
          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example.   Given a  function $\mathit{f}$,  we now  describe  how to
generate    equations    for     the    demand    transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
consider $\Lanv{1}{x_i}$,  the  liveness of    the
$i^{\mbox{\footnotesize  th}}$
parameter $x_i$ at the  program point at the  beginning of
$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
  live-define}, this should be the
same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
\clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

@@

To avoid analyzing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls by a {\em summary demand}  and calculate the liveness at each GC
point inside the  function based on this summary  demand.  Even though
we  lose precision  due to  this, our  experiments show  that  this is
acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
 \\
\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\epsilon
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$)
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
\begin{eqnarray*}
  \var{\Lanv{1}{\xl}}  &\rightarrow& 
  \clazy\sigma_{\length} \mid \acdr \var{\Df {\length}{1}}{\clazy\sigma_{\length}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\sigma_{\length}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
discussion that the decision problem  that we are interested in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

>>>>

\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_9\!\!:\, \cred{(\LET\  \pa\  \leftarrow
(\abigfunction\ \acdr)}$ \IN
          \UNL{3} \!\!$\pi_{10}\!\!:\, \cred{(\LET\  \pb\  \leftarrow
(}$\cred{+}$\cred{\ \pa\ \acdr)}$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ \cred{(\LET\ \pc}\
$\cred{\leftarrow  (\CONS\ \pb\ \NIL)}$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
          $\cred{(\LET\ \pw}\  $\cred{\leftarrow  (\length\ \pc)}$ \IN
          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

Given  a  function  $\mathit{f}$,  we now  describe  how  to  generate
equations  for the  demand  transformation \Lfonly$_\mathit{f}$.   The
program in Figure~\ref{fig:mot-example2} serves  as a running example.
Starting   with    a   symbolic   demand   $\sigma$,    we   determine
\Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In  particular, we  consider
$\Lanv{1}{x_i}$,  the liveness  of  the $i^{\mbox{\footnotesize  th}}$
parameter   $x_i$  at   the  program   point  at   the  beginning   of
$e_{\mathit{f}}$   (assumed  to   be  $\pi_1$).   By  the   rule  {\sc
  live-define},      this     should      be      the     same      as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
\clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

@@

To avoid analyzing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls and calculate the liveness at each GC
point inside the  function based on this approximation.
The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
 \\
\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item For each function $\mathit{f}$, equations defining \Df{f}{i} for
  use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item  For  each  function   $\mathit{f}$  (including  $\mainpgm$  for
  $e_\mainpgm$)  an equation  defining liveness  at each  GC point  of
  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
\begin{eqnarray*}
  \var{\Lanv{1}{\xl}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
discussion that the decision problem  that we are interested in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}
